Simulating an Interplanetary Internet

### 2024-04-10 paper/paper.tex
changed title
remove appendix reference
update the paper and add missing stats

\title{Analyzing Network Transfer Protocols on a Simulated Interplanetary Internet}
\author{Cody Auch, Mark Lysack, Jacob Janzen}
\date{\today}

@@

In a traditional network a majority of the network traffic will occur over a wired
connection, especially for transfers of large amounts of data.
This means that the error is relatively low. This is no longer the
case in the case of satellites sending messages over potentially millions of
kilometers. Atmosphere in the Earth orbit, solar radiation, and dust cause
increased error in sending the message between entities. Inverse square law
dictates that over these large periods the energy needed to send the message so
that it can be read at its destination. Even the Doppler effect is needed to be
accounted for to tune into radio frequencies on the ISS. ~\cite{qslCompensatingDoppler} High error means that
sending a message can no longer be assumed to be sent reliably and has to be
compensated for. Beyond high error, large distances between entities and the
physical limitation that messages cannot travel faster than the speed of light.
Light, or any other energy takes approximately 8 minutes to get from the Sun to Earth, ~\cite{nasa_earth} and
thus for a round trip will take 16 minutes to complete a simple handshake, given
that the messages were sent correctly.

@@

Based on these simplifications, a physical simulation with several goals. To
calculate if a line of sight could be made between two communicating entities,
that is if a message can be sent, to calculate the error rate when sending a
message, and to calculate the message propagation delay for the message to be
sent. Line of sight was achieved using simple geometry calculations. (See
appendix) All entities are able to block signals if the ray passes through the
radius of the entity. One of the largest factors contributing to transmission
error is the distance traveled by the electromagnetic signal between
transmitting and receiving nodes. By generalizing Friis' transmission
formula\cite{Friis} for isotropic emitters and receivers, we can obtain the
ratio of signal intensity received to signal intensity transmitted, known as
Free Space Path Loss (FSPL). We can use FSPL as our transmission error rate,
giving the following expression: $Error Rate = {(\frac{4 \pi d}{\lambda})}^2$
where $d$ is the distance between nodes in metres and $\lambda$ is the
wavelength of the electromagnetic wave used in transmission. As much of space
communications utilizes the Ka-band of frequencies (27--40
GHz)\cite{Morabito_Hastrup} we opted to use a transmission frequency of 30Ghz.
The transmission wavelength can be then calculated according to
$\lambda = \frac{c}{f}$, where $c$ is the speed of light and $f$ is the
transmission frequency, giving a transmission wavelength of about 9.99~mm.
Propagation delay is a simple function of distance and the speed of light.

@@

% UDP had an average time from being queued to received of
% 195.4 seconds, an average time being queued of
% 18.36 seconds, and an average message time in transit
% 177.0 seconds.

The average amount of data sent was 
1051.6 bytes. The number of bytes 
Notice the scientific notation in the 
corner of the graph.

% remove this line if stats are re-calculated
The average time that a message was in transit was close to 216 seconds.
The time spent waiting in queue was 0.094\% of the time it took for a message to completely 
transmit.

The graph shows a large amount of 
sensitivity to the motion of the planets which can be 
seen by the curvature of the graph. At time around 28000
seconds, the topology suddenly changed. Caused by a satellite
blocking the path between Mars and Earth. This caused a secondary, 
less efficient path to be taken. This shows why there are several 
"layers" of time which it took. They likely represent the different paths 
which the message took before being sent to its destination. Each quanta in
the time it took to complete the message 
represents the smaller jump it takes to get from a satellite to its 
orbiting entity. The increase in time likely represents that while the 
direct Earth and Mars link was down the smaller satellites were actually orbiting 
away from the message destination. UDP messages spent much more time being queued than 
the other messages. This may imply that messages were being transferred back and forth 
between two closely orbiting objects, which could explain the many different layers 
which it can take for a message to be sent. It could also be that since the average 
message is larger than TCP or New Reno it takes longer for the message to be sent 
and it is transmitting.

@@

The average amount of bytes transferred is 394 bytes. This is likely from 
the difference in the amounts of bytes transferred in an ACK message compared 
to when data is being sent.

Unlike in UDP the motion of the planets is not observable in the graph. However this is 
because there are several different paths which the TCP message could have taken which can be seen. 
Since TCP requires a higher amount of base time to establish a connection compared to UDP, this 
means that when it "hops" between entities it adds a larger base time.

@@

New Reno, a modified version of TCP used to help elevate congestion ~\cite{Nahar2016}, was used 
in place of SCP TP. New Reno behaved very similar to TCP. Its average data 
transfer was 407.5 bytes.

@@

Since a large amount of propagation delay dominates the amount of time that it takes for messages total 
transmit time from being queued to being received at the appropriate node, other increases in performance 
mean only a slight decrease in time sending the message. An important statistic to pay attention to is that 
UDP almost always sent a message with a size of 1052 bytes. This means that even though TCP, New Reno, and 
UDP had sent a similar number of messages, the actual throughput for UDP was significantly higher, which 
is much more significant in an IPN situation. In the data graphs we can see that not only was the maximum 
data being sent was less than UDP, but also had to send messages of very little size, most likely acknowledgements,
as well. New Reno was seen to improve on the queuing time compared to both TCP and UDP. Which would mean in a 
high reliability situation and in situations where higher traffic is sent through a network with a relatively low 
bandwidth, like a IPN, where satellites are limited by how much data they can transmit, New Reno's effect can be 
more significant. This should only be used in a connection where propagation delay is less of an issue like 
between planet and satellite communication. Queuing time in UDP took a more significant percentage of its 
time than both TCP and New Reno. In general however, UDP is the best option for primary communication between 
Mars and Earth communication.

@@

The physics engine could be expanded to included 3D orbits, particularly around planets, where networks of 
satellites could be created to create a proper level of communication that is maintained today, with geocentric 
and polar orbiting satellites. Connection points from a planet to off the planet could have a direction which it 
needs to send a message in for it to be able to connect. For example a rover on Mars in our simulation could technically 
send a message in any direction, even if they were sending it through the surface of the planet.The error model in 
the physics engine could be expanded to include interference like solar radiation, atmosphere, and other interference.

@@

\printbibliography

>>>>

\title{Simulating an Interplanetary Internet}
\author{Cody Auch, Mark Lysack, Jacob Janzen}
\date{\today}

@@

In a traditional network a majority of the network traffic will occur over a
wired connection, especially for transfers of large amounts of data. This means
that the error is relatively low. This is no longer the case in the case of
satellites sending messages over potentially millions of kilometers. Atmosphere
in the Earth orbit, solar radiation, and dust cause increased error in sending
the message between entities. Inverse square law dictates that over these large
periods the energy needed to send the message so that it can be read at its
destination. Even the Doppler effect is needed to be accounted for to tune into
radio frequencies on the ISS.~\cite{qslCompensatingDoppler} High error means
that sending a message can no longer be assumed to be sent reliably and has to
be compensated for. Beyond high error, large distances between entities and the
physical limitation that messages cannot travel faster than the speed of light.
Light, or any other energy takes approximately 8 minutes to get from the Sun to
Earth,~\cite{nasa_earth} and thus for a round trip will take 16 minutes to
complete a simple handshake, given that the messages were sent correctly.

@@

Based on these simplifications, a physical simulation with several goals. To
calculate if a line of sight could be made between two communicating entities,
that is if a message can be sent, to calculate the error rate when sending a
message, and to calculate the message propagation delay for the message to be
sent. Line of sight was achieved using simple geometry calculations. All
entities are able to block signals if the ray passes through the radius of the
entity. One of the largest factors contributing to transmission error is the
distance traveled by the electromagnetic signal between transmitting and
receiving nodes. By generalizing Friis' transmission formula\cite{Friis} for
isotropic emitters and receivers, we can obtain the ratio of signal intensity
received to signal intensity transmitted, known as Free Space Path Loss (FSPL).
We can use FSPL as our transmission error rate, giving the following expression:
$Error Rate = {(\frac{4 \pi d}{\lambda})}^2$ where $d$ is the distance between
nodes in metres and $\lambda$ is the wavelength of the electromagnetic wave used
in transmission. As much of space communications utilizes the Ka-band of
frequencies (27--40 GHz)\cite{Morabito_Hastrup} we opted to use a transmission
frequency of 30Ghz. The transmission wavelength can be then calculated according
to $\lambda = \frac{c}{f}$, where $c$ is the speed of light and $f$ is the
transmission frequency, giving a transmission wavelength of about 9.99~mm.
Propagation delay is a simple function of distance and the speed of light.

@@

UDP had an average time from being queued to received of 283.64 seconds, an
average time being queued of 21.92 seconds, and an average message time in
transit 261.72 seconds. The average amount of data sent was 1052.0 bytes. Notice
the scientific notation in the corner of the graph.

The graph shows a large amount of sensitivity to the motion of the planets which
can be seen by the curvature of the graph. At time around 28000 seconds, the
topology suddenly changed. Caused by a satellite blocking the path between Mars
and Earth. This caused a secondary, less efficient path to be taken. This shows
why there are several ``layers'' of time which it took. They likely represent the
different paths which the message took before being sent to its destination.
Each quanta in the time it took to complete the message represents the smaller
jump it takes to get from a satellite to its orbiting entity. The increase in
time likely represents that while the direct Earth and Mars link was down the
smaller satellites were actually orbiting away from the message destination. UDP
messages spent much more time being queued than the other messages. This may
imply that messages were being transferred back and forth between two closely
orbiting objects, which could explain the many different layers which it can
take for a message to be sent. It could also be that since the average message
is larger than TCP or New Reno it takes longer for the message to be sent and it
is transmitting.

@@

The average amount of bytes transferred is 320.44 bytes. This is likely from the
difference in the amounts of bytes transferred in an ACK message compared to
when data is being sent.

Unlike in UDP the motion of the planets is not observable in the graph. However
this is because there are several different paths which the TCP message could
have taken which can be seen. Since TCP requires a higher amount of base time to
establish a connection compared to UDP, this means that when it ``hops'' between
entities it adds a larger base time.

@@

New Reno, a modified version of TCP used to help reduce
congestion~\cite{Nahar2016}, was used in place of SCP TP. New Reno behaved very
similar to TCP. Its average data transfer was 307.18 bytes.

@@

Since a large amount of propagation delay dominates the amount of time that it
takes for messages total transmit time from being queued to being received at
the appropriate node, other increases in performance mean only a slight decrease
in time sending the message. An important statistic to pay attention to is that
UDP almost always sent a message with a size of 1052 bytes. This means that even
though TCP, New Reno, and UDP had sent a similar number of messages, the actual
throughput for UDP was significantly higher, which is much more significant in
an IPN situation. In the data graphs we can see that not only was the maximum
data being sent was less than UDP, but also had to send messages of very little
size, most likely acknowledgments, as well. New Reno was seen to improve on the
queuing time compared to both TCP and UDP. Which would mean in a high
reliability situation and in situations where higher traffic is sent through a
network with a relatively low bandwidth, like a IPN, where satellites are
limited by how much data they can transmit, New Reno's effect can be more
significant. This should only be used in a connection where propagation delay is
less of an issue like between planet and satellite communication. Queuing time
in UDP took a more significant percentage of its time than both TCP and New
Reno. In general however, UDP is the best option for primary communication
between Mars and Earth communication.

@@

The physics engine could be expanded to included 3-D orbits, particularly around
planets, where networks of satellites could be created to create a proper level
of communication that is maintained today, with geocentric and polar orbiting
satellites. Connection points from a planet to off the planet could have a
direction which it needs to send a message in for it to be able to connect. For
example a rover on Mars in our simulation could technically send a message in
any direction, even if they were sending it through the surface of the
planet.The error model in the physics engine could be expanded to include
interference like solar radiation, atmosphere, and other interference.

@@

\printbibliography{}
